---
title: "Exploratory Data Analysis on TTC Transit Delay Data - Potential Correlations Between Specific Time Periods and Delay Frequencies"
author: "Zhaotian Li, Yonghao Li, Hongtianxu Hua"
date: "8 Feb, 2020"
output: pdf_document
---

## 1. Abstract

  This paper is concerned with the potential effects of different times of day/week/year on the delay frequencies of the TTC transit system, including subway, buses, and streetcars. In particular, it uses exploratory data analytic methods (summary statistics, grouping, and plotting) on the TTC Delay Datasets (from Open Data Toronto) to clearly show that the number of delays peaks in the morning and afternoon rush hours. More importantly, we’ve found that “Injured or ill Customer”, “Speed Control”, and “Passenger Assistance Alarm Activated - No Trouble Found” are the most influential reasons for delay occurrences. Our initial hypothesis was that the frequency of daily/weekly/monthly delays should remain relatively consistent across time. However, by doing exploratory data analysis, we found that delay occurrences correlate with specific time periods. A predictive model built from this EDA might be able to predict the frequency of delays by temporal factors. This paper will give insights to the factors contributing to TTC delays and might be of interest to the authorities and management in their efforts to improve transit efficiency.

## 2. Introduction

Toronto Transit Commission (TTC) is the official public transport agency that manages all public transits within the Toronto area. The TTC network includes buses, streetcars, and subways, all of which are crucially important to Torontonians’ daily commute. Despite billions of dollars worth of operating budget (TTC and Wheel-Trans Operating Budgets, 2020) and revenue, the TTC often disappoints commuters. Such dissatisfaction has been studied and reported, with Toronto being ranked “worst city for commuting in North America” (GlobalNews.ca, 2018). One obvious issue that we explored is the delay in commute time. In a survey conducted by Forum Poll in May 2019, over half of the collected responses indicated that their travel on the TTC is often impacted by unexpected delays.

By studying the TTC delay datasets, we hope to figure out potential factors impacting the smooth operation of transit systems, which might be helpful in figuring out ways to improve passenger satisfaction as well as increase operating efficiency. Identifying the key factors can provide insights for TTC management to tackle the delay issue and plan for future improvements.


```{r, setup, include=FALSE}

library(opendatatoronto)
library(tidyverse)
library(stringr)
library(skimr)
library(janitor)
library(readr)
library(lubridate)
library(psych)
library(kableExtra)

############################################
# Fetch subway data
############################################
subway_delay_id <- "996cfe8d-fb35-40ce-b569-698d51fc683b"

res <- list_package_resources(subway_delay_id)
res <- res %>% mutate(year = str_extract(name, "201.?"))
delay_2019_ids <- res %>% filter(year==2019) %>% select(id) %>% pull()

delay_2019 <- c()
for(i in 1:length(delay_2019_ids)) {
  delay_2019 <- bind_rows(delay_2019, get_resource(delay_2019_ids[i]))
}

# make the column names nicer to work with
delay_2019 <- clean_names(delay_2019)

# fetch code books
delay_codes <- get_resource("fece136b-224a-412a-b191-8d31eb00491e")
delay_data_codebook <- get_resource("54247e39-5a7d-40db-a137-82b2a9ab0708")

# Replace code with meaningful text
delay_2019 <- delay_2019 %>% 
  left_join(delay_codes %>% rename(code = `SUB RMENU CODE`, 
                                   code_desc = `CODE DESCRIPTION...3`) 
            %>% select(code, code_desc)) 

delay_2019 <- delay_2019 %>%
  mutate(code_srt = ifelse(line=="SRT", code, "NA")) %>% 
  left_join(delay_codes %>% rename(code_srt = `SRT RMENU CODE`, 
                                   code_desc_srt = `CODE DESCRIPTION...7`) 
            %>% select(code_srt, code_desc_srt))  %>% 
  mutate(code = ifelse(code_srt=="NA", code, code_srt),
         code_desc = ifelse(is.na(code_desc_srt), code_desc, code_desc_srt)) %>% 
  select(-code_srt, -code_desc_srt)

most_frequent_codes <- delay_2019 %>% 
    group_by(code_desc) %>% summarise(count = n()) %>%
    top_n(n = 3, wt = count) %>% .$code_desc

# this function convert time (Ex. 15:30) into standard format (Ex. 2001-01-01 15:30)
# then floor the formatted time into hour
# the conversion is necessary because round() function doesn't support non-standard time format
custom_floor_time <- function(time) {
  tt <- strptime(paste("2001-01-01", time), format="%Y-%m-%d %H:%M")
  return(format(round(tt, units="hours"), format="%H:%M"))
}

hourly_delay <- delay_2019 %>% 
    group_by(hour=custom_floor_time(time), code_desc) %>% summarise(count = n()) 

daily_delay <- delay_2019 %>% 
    group_by(day, code_desc) %>%
    summarise(count = n())

monthly_delay <- delay_2019 %>% 
    group_by(month=floor_date(date, "month"), code_desc) %>%
    summarise(count = n()) 

```

## 3. Exploratory Data Analysis

### 3.1 Sanity Check

Check days of week
```{r, echo=FALSE}
unique(delay_2019$day)
```

We are mostly interested in the distribution of min_delay and min_gap (both contain number of minutes as data).
```{r, warning=FALSE, echo=FALSE}
describeBy(delay_2019 %>% select(min_delay, min_gap)) %>% kable() %>% kable_styling(latex_options = "scale_down") %>% footnote(general = "Table 1", general_title="", footnote_as_chunk = T)

```
Based on table 1, we can see that these two columns are positively skewed, with most values being short delays. This makes sense because subway delays are mostly short but frequent. Therefore, we can ensure the logic of our most important columns is correct.

### 3.2 Duplication
```{r, message=FALSE, echo=FALSE}
head(get_dupes(delay_2019), 2)  %>% kable() %>% kable_styling(latex_options = "scale_down") %>% footnote(general = "Table 2", general_title="", footnote_as_chunk = T)
```
In table 2, most of the duplications are caused by missing code and line value, and there're `r nrow(get_dupes(delay_2019))` of them out of `r nrow(delay_2019)` total data entries. Thus we choose to remove them before our analysis.
```{r, echo=FALSE}
delay_2019 <- delay_2019 %>% distinct()
```

### 3.2 Missing values
```{r, message=FALSE, warning=FALSE, echo=FALSE}
delay_2019 %>% 
  summarise_all(.funs = funs(sum(is.na(.))/nrow(delay_2019)*100)) %>% kable() %>% kable_styling(latex_options = "scale_down") %>% footnote(general = "Table 3", general_title="", footnote_as_chunk = T)
```
From table 3, most of the missing values exist in *bound* and *line* columns, but they are not the main factors in our analysis.

### 3.4 Visualizing Distributions

We started our analysis by finding relationships between hours and amount of delay incidents.


**A. Hourly pattern**

**A.1**
Within 24 hours of a day, when did TTC delay occur most frequently?

```{r, fig.width=8, fig.height=3, echo=FALSE}
hourly_delay %>%
  group_by(hour) %>% summarise(count = sum(count)) %>%
  ggplot(aes(x = hour, y = count)) + geom_bar(stat = "identity") +
  labs(caption="Figure 1") +
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Figure 1 shows the delay frequency counts grouped by each hour of the day, we can see that delay frequencies peak at around 8-10 am, 5-7 pm, and 11 pm. We also interested in seeing what role the variable code could play in hourly data.


**A.2**

In table 4, we selected 3 most frequent code that causes delay in 2019 and applied them in the following analysis.

```{r, echo=FALSE}
delay_2019 %>% 
    group_by(code_desc) %>% summarise(count = n()) %>%
    top_n(n = 3, wt = count)  %>% kable() %>% kable_styling() %>% footnote(general = "Table 4", general_title="", footnote_as_chunk = T)
```

Within 24 hours of a day, when did TTC delay occur most frequently and for what kind of reason? We chose the top 3 most frequent code of delay in 2019.

```{r, fig.width=8, fig.height=3, echo=FALSE}
hourly_delay %>%
  filter(code_desc %in% most_frequent_codes) %>%
  mutate(pct = paste0(round(count / sum(count), 2) * 100, "%")) %>%
  ggplot(aes(hour, count, fill=code_desc)) +
  geom_bar(position="stack", stat="identity") +
  geom_text(aes(label = ifelse(count > 15, pct, '')), 
          size = 3, position = position_stack(vjust = 0.5)) +
  labs(caption="Figure 2") +
  theme(legend.position="bottom", 
        legend.text=element_text(size=7.5), 
        axis.text.x = element_text(angle = 45, hjust = 1))
```


**B. Daily pattern**

Within 7 days of a week, when did TTC delay occur most frequently and for what kind of reason?

```{r, fig.width=8, fig.height=3, echo=FALSE}
# stacked bar
daily_delay %>%
  filter(code_desc %in% most_frequent_codes) %>%
  mutate(pct = paste0(round(count / sum(count), 2) * 100, "%")) %>%
  ggplot(aes(day, count, fill=code_desc)) + geom_bar(position="stack", stat="identity") +
  geom_text(aes(label = pct), size = 3, position = position_stack(vjust = 0.5)) + 
  labs(caption="Figure 3") +
  theme(legend.position="bottom", 
        legend.text=element_text(size=7.5))
```

**C. Monthly pattern**

Within 12 month of a year, when did TTC delay occur most frequently and for what kind of reason?

```{r, fig.width=8, fig.height=3, echo=FALSE}
plt_c2 <- monthly_delay %>%
  group_by(month, code_desc) %>%
  summarise(count = sum(count)) %>%
  filter(code_desc %in% most_frequent_codes) %>%
  group_by(month) %>%
  mutate(pct = paste0(round(count / sum(count), 2) * 100, "%"))

# stacked bar
plt_c2 %>%
  ggplot(aes(month, count, fill=code_desc)) +
  geom_bar(position="stack", stat="identity") +
  geom_text(aes(label = pct), 
          size = 3, position = position_stack(vjust = 0.5)) +
  labs(caption="Figure 4") +
  theme(legend.position="bottom", 
        legend.text=element_text(size=7.5))
```

**Research Question**

A predictive model built from this EDA might be able to predict the frequency of delays by temporal factors. In further investigation using linear regression models, we will consider the following factors:
Potential predictor variables: hour of the day, day of the week, month of the year, subway line
Response variable: delay time (in particular, what kind of delay might most likely occur given a date, time and subway line).


### Ethical issues

While we look into the objective factors like date and time, we have no intention of neglecting people working at TTC. They make great contributions to keeping this huge public transportation system running. On the other hand, this study is an early stage of data exploration with no intention of ascribing delays to any individual or group behavior. It’s more of a descriptive study than finding causal relationships.

Looking into the delay patterns is only for a research purpose; the study does not encourage readers to avoid or disdain public transportation. This is just an analysis into historic numbers that might help TTC with trouble-shooting; any possible prediction models and results based on the study should not be treated as the real situation at TTC.

### Shortcomings or weaknesses

The current datasets and variables may not be adequate to identify the most significant predictor variables and generate a better regression model in the next step. 

Other variables can be introduced to this dataset like weather, temperature, event calendar, logs of other accidents happened in the city, etc.

There might be even more variables that are not easily accessible, e.g. condition of the subway cars and tracks, employee schedule and shifts, the functionality of signals, etc. In order to solve the problem of delay, in-depth research and first-hand datasets are indispensable. 

\newpage
# References

Sharla Gelfand (2019). opendatatoronto: Access the City of Toronto Open Data Portal. R package version 0.1.1. https://CRAN.R-project.org/package=opendatatoronto

Wickham et al., (2019). Welcome to the tidyverse. Journal of Open Source Software, 4(43), 1686. https://doi.org/10.21105/joss.01686

Hadley Wickham (2019). stringr: Simple, Consistent Wrappers for Common String Operations. R package version 1.4.0. https://CRAN.R-project.org/package=stringr

Elin Waring, Michael Quinn, Amelia McNamara, Eduardo Arino de la Rubia, Hao Zhu and Shannon Ellis (2019). Skimr: Compact and Flexible Summaries of Data. R package version 2.0.2. https://CRAN.R-project.org/package=skimr

Sam Firke (2019). janitor: Simple Tools for Examining and Cleaning Dirty Data. R package version 1.2.0. https://CRAN.R-project.org/package=janitor

Hadley Wickham, Jim Hester and Romain Francois (2018). readr: Read Rectangular Text Data. R package version 1.3.1. https://CRAN.R-project.org/package=readr

Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with lubridate. Journal of Statistical Software, 40(3), 1-25. http://www.jstatsoft.org/v40/i03/.

Revelle, W. (2019) psych: Procedures for Personality and Psychological Research, Northwestern University, Evanston, Illinois, USA. https://CRAN.R-project.org/package=psych Version = 1.9.12

Hao Zhu (2019). kableExtra: Construct Complex Table with 'kable' and Pipe Syntax. R package version 1.1.0. https://CRAN.R-project.org/package=kableExtra

2020 TTC and Wheel-Trans Operating Budgets. ( 2019, December 16). https://www.ttc.ca/About_the_TTC/Commission_reports_and_information/Commission_meetings/2019/December_16/Reports/1_2020_TTC_and_Wheel_Trans_Operating_Budgets.pdf

Global News. (2018, June 21). Toronto ranked worst city for commuting in North America: study. https://globalnews.ca/news/4287922/toronto-worst-city-commuting-north-america/

City News. (2018, June 20). Commute times longer now than 6 years ago: poll. https://toronto.citynews.ca/2019/06/20/toronto-commute-times-poll/

   
\newpage

# Appendix

```{r, eval=FALSE}
library(opendatatoronto)
library(tidyverse)
library(stringr)
library(skimr)
library(janitor)
library(readr)
library(lubridate)
library(psych)
library(kableExtra)

############################################
# Fetch subway data
############################################
subway_delay_id <- "996cfe8d-fb35-40ce-b569-698d51fc683b"

res <- list_package_resources(subway_delay_id)
res <- res %>% mutate(year = str_extract(name, "201.?"))
delay_2019_ids <- res %>% filter(year==2019) %>% select(id) %>% pull()

delay_2019 <- c()
for(i in 1:length(delay_2019_ids)) {
  delay_2019 <- bind_rows(delay_2019, get_resource(delay_2019_ids[i]))
}

# make the column names nicer to work with
delay_2019 <- clean_names(delay_2019)

# fetch code books
delay_codes <- get_resource("fece136b-224a-412a-b191-8d31eb00491e")
delay_data_codebook <- get_resource("54247e39-5a7d-40db-a137-82b2a9ab0708")

############################################
# Merge code book with subway data
############################################
# Replace code with meaningful text
delay_2019 <- delay_2019 %>% 
  left_join(delay_codes %>% rename(code = `SUB RMENU CODE`, 
                                   code_desc = `CODE DESCRIPTION...3`) 
            %>% select(code, code_desc)) 

delay_2019 <- delay_2019 %>%
  mutate(code_srt = ifelse(line=="SRT", code, "NA")) %>% 
  left_join(delay_codes %>% rename(code_srt = `SRT RMENU CODE`, 
                                   code_desc_srt = `CODE DESCRIPTION...7`) 
            %>% select(code_srt, code_desc_srt))  %>% 
  mutate(code = ifelse(code_srt=="NA", code, code_srt),
         code_desc = ifelse(is.na(code_desc_srt), code_desc, code_desc_srt)) %>% 
  select(-code_srt, -code_desc_srt)

most_frequent_codes <- delay_2019 %>% 
    group_by(code_desc) %>% summarise(count = n()) %>%
    top_n(n = 3, wt = count) %>% .$code_desc

############################################
# EDA: checking data quality
############################################
# check days for the data
unique(delay_2019$day)

# check statistics of min_delay and min_gap
describeBy(delay_2019 %>% select(min_delay, min_gap)) %>% 
  kable() %>% 
  kable_styling(latex_options = "scale_down") %>%
  footnote(general = "Table 1", general_title="", footnote_as_chunk = T)

# check duplications
head(get_dupes(delay_2019), 2)  %>% 
  kable() %>% 
  kable_styling(latex_options = "scale_down") %>% 
  footnote(general = "Table 2", general_title="", footnote_as_chunk = T)


# remove duplications
delay_2019 <- delay_2019 %>% distinct()

# check missing values
delay_2019 %>% 
  summarise_all(.funs = funs(sum(is.na(.))/nrow(delay_2019)*100)) %>% 
  kable() %>%  
  kable_styling(latex_options = "scale_down") %>% 
  footnote(general = "Table 3", general_title="", footnote_as_chunk = T)

############################################
# EDA: data preparation
############################################
# this function convert time (Ex. 15:30) into standard format (Ex. 2001-01-01 15:30)
# then floor the formatted time into hour
# the conversion is necessary because round() function 
# doesn't support non-standard time format
custom_floor_time <- function(time) {
  tt <- strptime(paste("2001-01-01", time), format="%Y-%m-%d %H:%M")
  return(format(round(tt, units="hours"), format="%H:%M"))
}

hourly_delay <- delay_2019 %>% 
    group_by(hour=custom_floor_time(time), code_desc) %>% summarise(count = n()) 

daily_delay <- delay_2019 %>% 
    group_by(day, code_desc) %>%
    summarise(count = n())

monthly_delay <- delay_2019 %>% 
    group_by(month=floor_date(date, "month"), code_desc) %>%
    summarise(count = n()) 

############################################
# EDA: plotting section A hourly pattern
############################################

# Figure 1 
# Within 24 hours of a day, when did TTC delay occur most frequently?
hourly_delay %>%
  group_by(hour) %>% summarise(count = sum(count)) %>%
  ggplot(aes(x = hour, y = count)) + geom_bar(stat = "identity") +
  labs(caption="Figure 1") +
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Table 4
# 3 most frequent code that causes delay in 2019
delay_2019 %>% 
    group_by(code_desc) %>% summarise(count = n()) %>%
    top_n(n = 3, wt = count)  %>%
    kable() %>% kable_styling() %>%
    footnote(general = "Table 4", general_title="", footnote_as_chunk = T)

# Figure 2
# Within 24 hours of a day, when did TTC delay occur most frequently 
# and for what kind of reason? We chose the top 3 most frequent code of delay in 2019.
hourly_delay %>%
  filter(code_desc %in% most_frequent_codes) %>%
  mutate(pct = paste0(round(count / sum(count), 2) * 100, "%")) %>%
  ggplot(aes(hour, count, fill=code_desc)) +
  geom_bar(position="stack", stat="identity") +
  geom_text(aes(label = ifelse(count > 15, pct, '')), 
          size = 3, position = position_stack(vjust = 0.5)) +
  labs(caption="Figure 2") +
  theme(legend.position="bottom", 
        legend.text=element_text(size=7.5), 
        axis.text.x = element_text(angle = 45, hjust = 1))

############################################
# EDA: plotting section B daily pattern
############################################
# Figure 3
# Within 7 days of a week, when did TTC delay occur most frequently and
# for what kind of reason?
daily_delay %>%
  filter(code_desc %in% most_frequent_codes) %>%
  mutate(pct = paste0(round(count / sum(count), 2) * 100, "%")) %>%
  ggplot(aes(day, count, fill=code_desc)) + geom_bar(position="stack", stat="identity") +
  geom_text(aes(label = pct), size = 3, position = position_stack(vjust = 0.5)) + 
  labs(caption="Figure 3") +
  theme(legend.position="bottom", 
        legend.text=element_text(size=7.5))

############################################
# EDA: plotting section C month pattern
############################################
# shape data to plot monthly pattern
plt_c2 <- monthly_delay %>%
  group_by(month, code_desc) %>%
  summarise(count = sum(count)) %>%
  filter(code_desc %in% most_frequent_codes) %>%
  group_by(month) %>%
  mutate(pct = paste0(round(count / sum(count), 2) * 100, "%"))

# Figure 4
# Within 12 month of a year, when did TTC delay occur most frequently 
# and for what kind of reason?
plt_c2 %>%
  ggplot(aes(month, count, fill=code_desc)) +
  geom_bar(position="stack", stat="identity") +
  geom_text(aes(label = pct), 
          size = 3, position = position_stack(vjust = 0.5)) +
  labs(caption="Figure 4") +
  theme(legend.position="bottom", 
        legend.text=element_text(size=7.5))

```


